---
title: "Manuscript"
output: 
  html_notebook:
    toc: true
    toc_depth: 4
    theme: lumen
date: '`r format(Sys.time(), "%B %d, %Y")`'
---

This R Markdown Notebook provides R codes to fit the extended cosinor model, fPCA and HHMM discussed in the manuscript XXX.

```{r}
# set up R environment
suppressPackageStartupMessages({
  library(tidyverse); library(scales);library(ggplot2); library(depmixS4)
  })
```

# Sample Data

In this sample actigraphy data, we provide two example subjects from NHANES 2011-2014. We use these two example subjects to illustrate the usage of R codes developed for three models and investigate their Rest-Activity Rhythms (RAR).

```{r}
dat <- readRDS("two_example_subjects_raw_actigraphy.rds")
# subject id
id <- unique(dat$SEQN)
```

```{r}
my_theme <- function(){
  theme_bw(base_size = 16, base_family = "sans") %+replace%
  theme(
    plot.title = element_text(
      size = 13, face = "bold", hjust = 0, margin=margin(b = 8, unit = "pt")
      ), 
    plot.subtitle = element_text(
      size = 12, hjust = 0, margin=margin(b = 8, unit = "pt")
      ), 
    legend.title = element_blank(), 
    legend.text = element_text(size=16),                  
    axis.text = element_text(size=12), 
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(angle = 0, vjust = 1, hjust=1), 
    axis.text.y = element_text(size = 10),
    axis.title=element_text(size=12,face="bold"),
    strip.text.y = element_text(size = 12), legend.position="right"
    )}
```

# Extended Cosinor Model

Step 1: Data preparation. Use logarithm transform the data.

```{r}
# subject 1 used in the manuscript
example.id <- id[1]
df <- dat %>% filter (SEQN == example.id)

# get date and time variables 
time_zone <- lubridate::tz(df$datetime)
df$date <- as.Date(df$datetime, tz=time_zone)
df$time <- format(df$datetime,"%H:%M:%S")
#dates <- unique(df$date)
#df$day <- match(df$date, dates)
df$day <- as.numeric(df$day)
# clock time in minutes
df$tm <- lubridate::hour(df$datetime)*60 + lubridate::minute(df$datetime) + lubridate::second(df$datetime)/60
df$thrs <- lubridate::hour(df$datetime) + lubridate::minute(df$datetime)/60  + lubridate::second(df$datetime)/3600
df$td <- (df$day - 1) * 1440 + df$tm
df$n <- seq(1, nrow(df))

# logarithm transformation for actigraphy data
df$log.act <- log(df$activity_avg + 1)
```

Step 2. Fit the cosine model, which is the null model to be compared with the improved model fitting from extended cosinor model.

```{r}
# Specify staring values for initial cosine model
mMeanAct <- mean(df$log.act, na.rm = FALSE)
aStartAct = mMeanAct - min(df$log.act, na.rm = FALSE)
#start_cosine = as.data.frame(cbind(mMeanAct, aStartAct))

# Initial Cosine Model
# @amp: Amplitude
# @M: MESOR
# @phi: acrophase
cosine <- stats::nls(
  log.act ~ (amp + M) + amp*cos((tm/60 - phi)*pi/12)
  , data = df
  , start = list(amp = aStartAct, M = mMeanAct, phi = 12)
  , algorithm = "port"
  , control = list(warnOnly = T)
  , trace=F
  , lower=c(0, 0, -6), upper=c(Inf, Inf, 30)
  )

# Saving estimates of Initial Cosine Model to use as starting values for Fitted Extended Cosine Model
cos.coefs <- stats::coef(cosine)
```

Step 3. Fit the extended cosinor model.

```{r}
ex_cosine <- NULL; attempt <- 0

# fit the model until convergence reached
while( is.null(ex_cosine) &&  attempt <= 20 ) {
  attempt <- attempt + 1
  try(
    ex_cosine <- stats::nls(
      log.act ~ mn + amp*(
        exp(beta*(cos((tm/60 - phi)*pi/12)- alpha))
        /(1+exp(beta*(cos((tm/60 - phi)*pi/12) - alpha)))
        )
      , data = df
      # starting values are from the fitting cosine model
      , start = list(
        amp=cos.coefs[[1]], mn=cos.coefs[[2]], phi=cos.coefs[[3]], alpha=0, beta=2
        )
      , control = list(maxiter = 1000, warnOnly = T)
      , trace = F, algorithm = "port"
      , lower = c(0, 0, -6, -1, 0), upper=c(Inf, Inf, 30, 1, Inf))
    , silent=T
  )}
```

Step 4. Extract model estimates.

```{r}
# Coefficient estimates of extended cosine
final.ests = as.data.frame(t(stats::coef(ex_cosine)))

# Predicted Values & Residuals
exp_trans <- function(x){return(exp(x) - 1)}

# Fit predicted values based on observed non-missing data
df$predicted <- as.numeric(stats::predict(ex_cosine, data.frame( tm = df$tm ) )) 

# Predicted based on log(act + 1)
df$resid <- df$log.act - df$predicted # residuals of log(act + 1)

# Predicting based on Interpolated Data
obs.per.min <- df$tm[2] - df$tm[1]

# Create interpolated data
interp.td <- with(df, seq(min(td), max(td), by = obs.per.min)) 

# Predictions based on interpolated data
interp.pred <- as.data.frame(
  as.numeric(stats::predict(ex_cosine, newdata = list(tm = interp.td)))
  ) 
colnames(interp.pred) <- "interp.pred"
# number of minutes after first midnight
interp.pred$td <- interp.td 

# Exponentiation of predicted values and residuals
df$e_predicted <- exp_trans(df$predicted)
df$e_resid <- exp_trans(df$log.act) - df$e_predicted
interp.pred$e_interp.pred <- exp_trans(interp.pred$interp.pred)

# Merge interpolated data with original data (for plotting purposes, keep all NAs)
df_interp <- merge(df, interp.pred, by="td", all.y = TRUE)

# Calculating F Statistic
diff.mean.sq <- (df$predicted - mean(df$log.act))^2
RSS <- sum(diff.mean.sq)/4
MSE <- sum(df$resid^2)/(nrow(df)-5)
F_stat <- RSS/MSE

# Calculating Other RAR Parameters
alpha = final.ests$alpha
beta = final.ests$beta
acrophase = final.ests$phi
amp = exp(final.ests$amp)
mesor = exp(final.ests$mn + final.ests$amp/2)
log_amp = final.ests$amp
log_mesor = final.ests$mn + final.ests$amp/2
tLeft = final.ests$phi - acos(final.ests$alpha)/(2*pi/24)
tRight = final.ests$phi + acos(final.ests$alpha)/(2*pi/24)
ExtCos_params <- as.data.frame(
  cbind(example.id, alpha, beta, acrophase, amp, mesor, log_amp, log_mesor, tLeft, tRight, F_stat)
  )

print(ExtCos_params)
```


```{r}
# Plot on Log Scale
plot_title <- "Observed and Predicted Activity (with missingness)"
day_labels <- unique(df$day)
day_labels <- as.character(day_labels[stats::complete.cases(day_labels)])

ggplot2::ggplot(df, aes(x = td, y = log.act, group=1)) +
  ggplot2::geom_point(size=1, color="grey20", shape=1) + 
  my_theme() + 
  ggplot2::geom_line(
    data=df
    , aes(x = td, y = predicted, group=1, color = as.factor(is.na(day)))
    , linewidth = 1, show.legend = F
    ) +
  ggplot2::labs(
    title = paste0('ID=', example.id)
    , x = "Days of Observation"
    , y = "Log(activity + 1)"
    , col = "Missing Data"
    ) +
  labs(
    title = paste0('ID=', example.id)
    , subtitle = "anti-logistic-transformed cosine curve"
    , x = "Days of Observation"
    , y = "Log(Activity Counts + 1)"
    , col = "Missing Data"
    ) +
  ggplot2::scale_x_continuous(
    breaks = seq(min(df$td), max(df$td), by = 1440)
    , labels = day_labels
    ) 
```

# Functional Principal Component Analysis

# Harmonic Hidden Markov Model

Step 1. Fit Harmonic Hidden Markov Model.

```{r}
# Sampling frequency in hour unit. In our study, sf = 5min/60min = 1/12 
sf <- 1/12 
df$lag <- seq(0,(nrow(df)-1)) 
# circadian oscillators
df$sin_part <- sin(2*pi*df$lag*sf/24) 
df$cos_part <- cos(2*pi*df$lag*sf/24) 

df$y <- sqrt(df$activity_avg)

HMM_hmc <- NULL; attempt <- 0
y_hmc <- depmix(
  y ~ 1, data = df
  , transition = ~ sin_part + cos_part
  , nstates = 3
  , family = gaussian()
  , ntimes = nrow(df)
  )

### first try without any data padding
while( is.null(HMM_hmc) &&  attempt <= 20 ) {
  attempt <- attempt + 1
  try( HMM_hmc <- fit(y_hmc, verbose=F), silent = T)
} 
attempt <- 0 ### reset counter
```


```{r}
### OPTIONAL
### in case of failed convergence, add data padding
while( is.null(HMM_hmc) &&  attempt <= 20 ) {
  ### adding a bit value to the zeros
  df_padded <- df
  ind <- which(df_padded$y == 0); #length(ind)
  df_padded$y[ind] <- log(
    rbinom(length(ind), size = 1, prob = 0.5) + 1
    )  ### equal chance to be 0 or 1
  
  y_hmc <- depmix(
    y ~ 1
    , data = df_padded
    , transition = ~ sin_part + cos_part
    , nstates = 3
    , family = gaussian()
    , ntimes = nrow(df_padded)
    )
  try(HMM_hmc <- fit(y_hmc,verbose=F), silent = T)
}   
attempt <- 0 ### reset counter
```

Step 2. Extract model estimates.

```{r}
# Function developed by (Huang, 2018) to extract estimates of key parameters, for example, mean levels of three hidden activity states and circadian oscillated time-varying transition probabilities and state probabilities.

source("Harmonic_HMMs.r")

# Note: the output printed from Harmonic_HMMs() function is the model fitting results from using depmixS4::depmix
HMM_results_hmc <- Harmonic_HMMs(
  HMM = HMM_hmc
  , sin_part = df$sin_part
  , cos_part = df$cos_part
  )
```

```{r}
# extract the oscillated state probability 
circadian_states_prob <- HMM_results_hmc$circadian_states_prob 
# convert time variable to clock time
time <- as.POSIXct(df$time, format = "%H:%M:%S") 
# self-define the start hour of a day at 12 pm
hour_day_start <- 12 
# define a function to locate the first start hour of a day, since there are multiple days of records.
find_day_start <- function(hour_day_start,time){ 
  hour_time <- hour(time)
  min_time <- minute(time) 
  A <- which(hour_time == 12) 
  B <- which(min_time < 60*sf) 
  one_day_start <- A[min(which(A %in% B, arr.ind = TRUE))] 
  return(one_day_start) 
} 

one_day_start <- find_day_start(hour_day_start = hour_day_start, time = time) 
# we use 5-mins data (sf=1/12) so that one day has 24*12=288 points 
one_day_end <- one_day_start + (24/sf-1) 
# one day profile of three states probability
one_day_prob <- circadian_states_prob[one_day_start:one_day_end, ]

# inactive state probability 
p1 <- one_day_prob$state_1 
# moderately active state probability 
p2 <- one_day_prob$state_2 
# highly active state probability
p3 <- one_day_prob$state_3 
```

Step 3. Calculate RAR parameters using one 24-hour state probabilities.

```{r}
# amount of rest: duration of rest per day (hours) 
rest_amount <- 24*mean(p1, na.rm = T)

# absolute index to compute the gravity center (in the unit of hours) of p1
# the index is in 5-min epochs, same as the actigraphy data.
index <- seq(0, 24-1/12, 1/12)  
# center of rest (CRC) which corresponds to the gravity center of p1
# Note from Huang, 2018: the definition of only valid with suitably defined hour_day_start where the major resting window is of one 24 hours.
center_rest <- sum(p1*index/sum(p1, na.rm = T), na.rm = T) 

# center_rest is the position in index not the clock time. 
## #One can easily convert it to clock time via: 
index_clocktime <- seq(hour_day_start, hour_day_start + 24 - sf, sf) 
clocktime <- index_clocktime 
clocktime[index_clocktime>24]<-(index_clocktime-24)[index_clocktime>24] 
center_rest_clock <- clocktime[which.min(abs(index-center_rest))] 

#worst clock : complete lack of circadian rhythm where the probability of rest is constant and equal to rest_amount/24
worst_p1 <- rep(mean(p1, na.rm = T), 24/sf)

# perfect clock: major resting window has no interruptions 
find_perfect_p1 <- function(center_rest, rest_amount, index, sf){ 
  
  perfect_p1 <- rep(0, 24/sf) 
  
  t1 <- center_rest - rest_amount/2 
  t2 <- center_rest + rest_amount/2 
  
  perfect_t2<-which.min(abs(index-center_rest)) 
  perfect_t1<-which.min(abs(index-center_rest+rest_amount/2)) 
  perfect_t3<-which.min(abs(index-center_rest-rest_amount/2)) 
  
  perfect_p1[perfect_t1:perfect_t3] <- 1 
  
  if(t2>max(index)){offset<-round((t2-max(index))/sf);perfect_p1[1:offset]<-1} 
  if(t1<min(index)){offset<-round((min(index)-t1)/sf);perfect_p1[-offset+L,L]<-1} 
  
  return(perfect_p1) 
} 

perfect_p1 <- find_perfect_p1(center_rest, rest_amount, index, sf) 

# Rhythmic index (RI). 
RI <- (sum(p1[which(perfect_p1 > 0)])*sf/rest_amount-rest_amount/24)*24/(24- rest_amount) 

# the clock time of the highest IA probability 
MIPP <- clocktime[which(p1 == max(p1))]
# the clock time of the highest HA probability 
MAPP <- clocktime[which(p3 == max(p3))]

# Gaussian emission parameters of activity levels
state_means <- HMM_results_hmc$obs_params$mean
state_sds <- HMM_results_hmc$obs_params$sd
# the mean levels from the original output from depmixS4::depmix() results are not in increasing order from low to high. Reorder the states to determine the estimates for inactive, moderately active, and highly active activity states with mean levels from low to high.
reorder_index <- order(state_means)
# Mean levels of hidden activity states
IA_mean <- state_means[reorder_index[1]]; IA_sd <- state_sds[reorder_index[1]]
MA_mean <- state_means[reorder_index[2]]; MA_sd <- state_sds[reorder_index[2]]
HA_mean <- state_means[reorder_index[3]]; HA_sd <- state_sds[reorder_index[3]]

HHMM_params <- as.data.frame(
  cbind(example.id, RI, RA = rest_amount, CRC = center_rest_clock
        , MIPP, MAPP, IA_mean, IA_sd, MA_mean, MA_sd, HA_mean, HA_sd)
  )
print(HHMM_params)
```


```{r}
index_position <- seq(0, 24, 2)
# again the start hour of a day is 12 pm, which is also used for plotting the Rest-Activity profile
index_lable <- c('12','14','16','18','20','22', '0/24','2','4','6','8','10','12') 

par(fig = c(0, 1, 0, 1), new = F) 
source("figure_RA_profile_HHMM.r")
figure_day_profile(1, one_day_prob, index, index_position, index_lable)
```

